---
title: "Lab 4 - Individual 1"
author: "Ashley Sackpraseuth"
date: "10/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Data

```{r}
library(tidyverse)
library(readxl)

defense <- read_excel('/Users/asack/OneDrive/Documents/ds202_lab4/cyclonesFootball2019.xlsx', sheet='Defensive')
str(defense)

offense <- read_excel('/Users/asack/OneDrive/Documents/ds202_lab4/cyclonesFootball2019.xlsx', sheet='Offensive')
str(offense)

biography <- read_excel('/Users/asack/OneDrive/Documents/ds202_lab4/cyclonesFootball2019.xlsx', sheet='Biography')
str(biography)
```

## Part one: Cleaning data
The player names and opponent names are factors.
All offensive and defensive statistics are numerical. The Weight column in biogrpahy is also numeric. (Optional: Learn to mutate multiple columns systematically using the ?across function in `dplyr)
Change the Height column in biogrpahy into numeric. (Hint: Use a unit that works well)
```{r}
defClean <- defense %>%
              mutate(across(c("Name", "Opponent_Opponent"), as.factor)) %>%
              mutate_if(is.character, as.numeric, na.rm = TRUE)
str(defClean)
  
  

offClean <- offense %>%
              mutate(across(c("Name", "Opponent_Opponent"), as.factor)) %>%
              mutate_if(is.character, as.numeric, na.rm = TRUE)
str(offClean)

  

bioClean <- biography %>%
              mutate(Name = as.factor(Name)) %>%
              mutate(Height = gsub("-", ".", Height)) %>% #Height is converted to ft.in
              mutate(across(c("Weight", "Height"), as.numeric))
str(bioClean)

```


## Part two: Tidying

1.Reformat the `defClean` data frame into a tidy format using `pivot_longer`. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named `stat`.

```{r}
defClean <- defClean %>%
              pivot_longer(Tackles_Solo:Pass_PB, names_to = 'stat', values_to='score')

print(defClean)
```


2. Compare the distributions of the defensive statistics. What defensive skills are rare?

```{r}
ggplot(defClean, aes(x = score)) + geom_histogram() + facet_wrap(~stat, scales = 'free_y')
```
  
  Turnovers (Turnover_FF, Turnover_FR, Turnover_INT) and Tackles_Sack are rare defensive skills. 

3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.


```{r}
#x-axis solo tackles against Iowa State
#y-axis solo tackles against West Virginia

soloPerGame <- defClean %>% 
                group_by(Opponent_Opponent) %>%
                summarize(solo = sum(Tackles_Solo)) 


soloPerGame %>% 
  filter(Opponent_Opponent == 'Iowa' | Opponent_Opponent == 'West Virginia') %>%
  pivot_wider(names_from = 'stat', values_from = 'score') %>%
  ggplot(aes(x='Iowa', y = 'West Virginia')) geom_point() + facet_wrap(~Name)
```


4. Separate the `Hometown` column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)

```{r}
bioClean %>% 
  separate(Hometown, into = c("City", "State"), sep = ",")
```

5. How many players are there from each state?

```{r}
bioClean %>%
  separate(Hometown, into = c("City", "State"), sep = ",") %>%
  group_by(State) %>%
  summarize(numPlayers = count(Name))
```


## Part three: Joining data frames

1. Find which states do the ISU defensive and offensive players come from. Create an appropriate summary.

```{r}

```

2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?

```{r}

```


3. Which ISU player(s) made big progress over the last year? 


```{r}

```
