---
title: "Lab 4 - Team"
author: "Ashley Sackpraseuth, Adam Ford"
date: "10/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Data

```{r}
library(tidyverse)
library(readxl)

defense <- read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
str(defense)

offense <- read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
str(offense)

biography <- read_excel('cyclonesFootball2019.xlsx', sheet='Biography')
str(biography)
```

## Part one: Cleaning data
The player names and opponent names are factors.
All offensive and defensive statistics are numerical. The Weight column in biogrpahy is also numeric. (Optional: Learn to mutate multiple columns systematically using the ?across function in `dplyr)
Change the Height column in biogrpahy into numeric. (Hint: Use a unit that works well)
```{r}
defClean <- defense %>%
              mutate(across(c("Name", "Opponent_Opponent"), as.factor)) %>%
              mutate_if(is.character, as.numeric, na.rm = TRUE)
str(defClean)
  
  

offClean <- offense %>%
              mutate(across(c("Name", "Opponent_Opponent"), as.factor)) %>%
              mutate_if(is.character, as.numeric, na.rm = TRUE)
str(offClean)

  

bioClean <- biography %>%
              mutate(Name = as.factor(Name)) %>%
              mutate(Height = gsub("-", ".", Height)) %>% #Height is converted to ft.in
              mutate(across(c("Weight", "Height"), as.numeric))
str(bioClean)

```


## Part two: Tidying

1.Reformat the `defClean` data frame into a tidy format using `pivot_longer`. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named `stat`.

```{r}
defCleanWide <- defClean
defClean <- defClean %>%
              pivot_longer(Tackles_Solo:Pass_PB, names_to = 'stat', values_to='score')

print(defClean)
```


2. Compare the distributions of the defensive statistics. What defensive skills are rare?

```{r}
ggplot(defClean, aes(x = score)) + geom_histogram() + facet_wrap(~stat, scales = 'free_y')
```
  
  Turnovers (Turnover_FF, Turnover_FR, Turnover_INT) and Tackles_Sack are rare defensive skills. 

3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.


```{r}
soloTackles <- defClean %>%
               filter(Opponent_Opponent == 'Iowa' | Opponent_Opponent == 'Notre Dame', stat == 'Tackles_Solo') %>%
    pivot_wider(names_from = Opponent_Opponent, values_from= score)

ggplot(soloTackles, aes(x=Iowa, y=`Notre Dame`)) + geom_point() + xlab('Solo tackles in Iowa game') + ylab('Solo tackles in Notre Dame game')

```

  According to the plot, Iowa State had better defense in the Notre game, because the plot shows more points with 2 or more solo tackles.

4. Separate the `Hometown` column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)

```{r}
bioClean <- bioClean %>% 
  separate(Hometown, into = c("City", "State"), sep = ", ")
head(bioClean)
```

5. How many players are there from each state?

```{r}
bioClean %>%
  group_by(State) %>%
  summarize(numPlayers = n())
```


## Part three: Joining data frames

1. Find which states do the ISU defensive and offensive players come from. Create an appropriate summary.

```{r}
defWithHomes <- defClean %>%
  select(Name) %>%
  distinct() %>%
  left_join(bioClean, by = "Name")

defWithHomes %>% 
  group_by(State) %>%
  summarize(numPlayers = n()) %>%
  na.omit() %>%
  arrange(desc(numPlayers))

offWithHomes <- offClean %>%
  select(Name) %>%
  distinct() %>%
  left_join(bioClean, by = "Name")

offWithHomes %>% 
  group_by(State) %>%
  summarize(numPlayers = n()) %>%
  na.omit() %>%
  arrange(desc(numPlayers))

```

2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?

```{r}
offClean$YdsResponsibleFor <- rowSums(offClean[,c("Rushing_YDS", "Receiving_YDS", "Passing_YDS")], na.rm=TRUE)
offClean %>%
  group_by(Name) %>%
  summarize(AvgYdsResponsibleFor = mean(YdsResponsibleFor)) %>%
  arrange(desc(AvgYdsResponsibleFor))
```

  I added a conglomerate statistic for offensive players that is Yards Responsible For which consists of a sum of passing, rushing and receiving. Then averaged the stat for each player. It is obvious that Brock Purdy's performance has the greatest single impact on the Iowa State team.

3. Which ISU player(s) made big progress over the last year? 
```{r}
defense18 <- read_excel('cyclonesFootball2018.xlsx', sheet='Defensive')
defCleanWide18 <- defense18 %>%
              mutate(across(c("Name", "Opponent_Opponent"), as.factor)) %>%
              mutate_if(is.character, as.numeric, na.rm = TRUE)
defCleanWide19 <- defCleanWide

defCleanWide18 <- defCleanWide18 %>%
  group_by(Name) %>%
  select(-Opponent_Opponent) %>%
  summarize_all(list(mean), na.rm=TRUE) 

defCleanWide19 <- defCleanWide19 %>%
  group_by(Name) %>%
  select(-Opponent_Opponent) %>%
  summarize_all(list(mean), na.rm=TRUE) 


colnames(defCleanWide18) <- paste("Last", colnames(defCleanWide18), sep = "_")
colnames(defCleanWide19) <- paste("This", colnames(defCleanWide19), sep = "_")
DefCompare <- inner_join(defCleanWide18, defCleanWide19, by = c("Last_Name" = "This_Name"))
DefCompare$DiffTacklesSolo <- DefCompare$This_Tackles_Solo - DefCompare$Last_Tackles_Solo
DefCompare$DiffTacklesSack <- DefCompare$This_Tackles_Sack - DefCompare$Last_Tackles_Sack
DefCompare$DiffTacklesAsst <- DefCompare$This_Tackles_ASST - DefCompare$Last_Tackles_ASST

DefCompare %>%
  select(Last_Name, DiffTacklesSolo, DiffTacklesSack, DiffTacklesAsst) %>%
  arrange(desc(DiffTacklesSolo)) %>%
  head(1)

DefCompare %>%
  select(Last_Name, DiffTacklesSolo, DiffTacklesSack, DiffTacklesAsst) %>%
  arrange(desc(DiffTacklesSack)) %>%
  head(1)

DefCompare %>%
  select(Last_Name, DiffTacklesSolo, DiffTacklesSack, DiffTacklesAsst) %>%
  arrange(desc(DiffTacklesAsst)) %>%
  head(1)
```

  These result shows the difference of the average of each player's important stats last season versus this season. We see Anthony Johnson most improved in Solo tackles and O'Rien Vance most improved in Sacks and Assisted Tackles.
  
```{r}
offense18 <- read_excel('cyclonesFootball2018.xlsx', sheet='Offensive')
offCleanWide18 <- offense18 %>%
              mutate(across(c("Name", "Opponent_Opponent"), as.factor)) %>%
              mutate_if(is.character, as.numeric, na.rm = TRUE)
offCleanWide19 <- offClean

offCleanWide18 <- offCleanWide18 %>%
  group_by(Name) %>%
  select(-Opponent_Opponent) %>%
  summarize_all(list(mean), na.rm=TRUE) 

offCleanWide19 <- offCleanWide19 %>%
  group_by(Name) %>%
  select(-Opponent_Opponent) %>%
  summarize_all(list(mean), na.rm=TRUE) 

colnames(offCleanWide18) <- paste("Last", colnames(offCleanWide18), sep = "_")
colnames(offCleanWide19) <- paste("This", colnames(offCleanWide19), sep = "_")
OffCompare <- inner_join(offCleanWide18, offCleanWide19, by = c("Last_Name" = "This_Name"))
OffCompare$DiffRecYds <- OffCompare$This_Receiving_YDS - OffCompare$Last_Receiving_YDS
OffCompare$DiffPassingYds <- OffCompare$This_Passing_YDS - OffCompare$Last_Passing_YDS
OffCompare$DiffRushingYds <- OffCompare$This_Rushing_YDS - OffCompare$Last_Rushing_YDS

OffCompare %>%
  select(Last_Name, DiffRecYds, DiffPassingYds, DiffRushingYds) %>%
  arrange(desc(DiffRecYds)) %>%
  head(1)

OffCompare %>%
  select(Last_Name, DiffRecYds, DiffPassingYds, DiffRushingYds) %>%
  arrange(desc(DiffPassingYds)) %>%
  head(1)

OffCompare %>%
  select(Last_Name, DiffRecYds, DiffPassingYds, DiffRushingYds) %>%
  arrange(desc(DiffRushingYds)) %>%
  head(1)
```
  
  These results are for the offensive players. Some NaN exist in categories where players have not done such an action, however this will not impact finding the "improvers". In Receiving Yards, Charlie Kolar has improved the most since last season. Brock Purdy has improved the most in Passing Yards. And Johnnie Lang has improved the most in Rushing Yards.
